\section{Connected Data}



\subsection{Update Protocols}

The recorded lectures are provided \myhref{https://www.youtube.com/watch?v=-P4Qks9J3o0}{here (Introduction)} and \myhref{https://www.youtube.com/watch?v=B70b-uyNHpw}{here (Update Protocols)}.

\begin{ejercicio}
    What is an update protocol and how does it differ from stream compression?

    An update protocol is a method for efficiently transmitting updates from IoT devices. It focuses on reducing the number of measurements needed and transmitted, while stream compression focuses on reducing the size of the data being transmitted.
\end{ejercicio}

\begin{ejercicio}
    Explain the update protocol ``Periodic Reporting'' and describe a situation in which it works very well and a situation in which it does not work well. Which other update protocol can we use to solve the problem?

    Periodic Reporting is an update protocol where the IoT device sends updates at regular intervals (every $T$ seconds). The period $T$ is fixed following a worst-case perspective: if the maximum speed of the moving object is $v_{\max}$ and the maximum allowed error is $d_{\max}$, then:
\[
T = \frac{d_{\max}}{v_{\max}}
\]
    \begin{itemize}
        \item This protocol works well when the object is moving at a constant speed next to the maximum speed, and it is moving in a straight line. In this case, the updates will be sent at the right time and the error will be within the allowed limit.
        \item This protocol does not work well when the object is moving at a very low speed or is stationary. In this case, the device will send updates unnecessarily, which can lead to wasted energy and bandwidth. To solve this problem, we can use any other update protocol that is more adaptive to the object's movement, such as the Dead Reckoning protocol, which sends updates only when the object deviates from its predicted path by more than a certain threshold.
    \end{itemize}
\end{ejercicio}

\newpage\subsection{Data Models}

The recorded lecture is provided \myhref{https://www.youtube.com/watch?v=ZZ3H8QTZKv8}{here}.

\begin{ejercicio}
    Give an example for an information that a Key-Value Model cannot represent but a Linked-data Based Model can.\\

    For example, a Key-Value Model cannot represent the following values:
    \begin{itemize}
        \item If you want to specify the unit of a measurement, you cannot do it in a Key-Value Model. For example, if you want to specify that the temperature is 25 degrees Celsius, you cannot represent the unit in a Key-Value Model. You could represent it as \texttt{(temperature, 25)} but you cannot specify that the unit is Celsius.
    \end{itemize}
\end{ejercicio}

\newpage\subsection{Stream Processing and Complex Event Processing}

The recorded lectures are provided \myhref{https://www.youtube.com/watch?v=lAdIGfW8fUg&feature=youtu.be}{here (Introduction)}, \myhref{https://www.youtube.com/watch?v=3oDF0BbPR0A&feature=youtu.be}{here (Stream Processing)} and \myhref{https://www.youtube.com/watch?v=8hbLKM3zH9w}{here (Complex Event Processing)}.

\begin{ejercicio}
    Compare Stream Processing with Complex Event Processing by discussing the different information models, what they each focus on and where they come from.

    \begin{itemize}
        \item \ul{Information Models}: Stream Processing understand data as a data stream (mesurement after measurement), while Complex Event Processing understand data as a sequence of events, where an event can be a measurement, but also a more complex event that is derived from other events.
        \item \ul{Focus}: Stream Processing focuses on processing data in real-time as it arrives, while Complex Event Processing focuses on detecting complex patterns and relationships between events, which may not be possible to detect in real-time.
        \item \ul{Origin}: Stream Processing comes from the field of data management and databases, while Complex Event Processing comes from the idea of publish-subscribe systems and event-driven architectures.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    What is a continuous query? Give an example for one in pseudo code (i.e., define your own language for continuous queries and provide an example with it). Include typical features like, e.g., a window(-ing) concept into your language.

    A continuous query is a query that runs continuously and processes data as it arrives. It is used in stream processing systems to analyze data in real-time. An example of a continuous query in pseudo code could be the one shown in Listings~\ref{lst:continuous_query}.
\begin{listing}
\begin{minted}{sql}
SELECT AVG(temperature)
FROM SensorData
WHERE location = 'living_room'
WINDOW TUMBLING (SIZE 1 MINUTE)
\end{minted}
\caption{Example of a continuous query that calculates the average temperature in the living room every minute.}
\label{lst:continuous_query}
\end{listing}
\end{ejercicio}

\begin{ejercicio}
    What are Event-Condition-Action rules and how do they differ from continuous queries? What is similar?

    Event-Condition-Action (ECA) rules are a type of rule that specifies an action to be taken when a certain event occurs and a certain condition is met.
    \begin{itemize}
        \item \ul{Difference}: ECA rules are more focused on triggering actions based on events and conditions, while continuous queries are more focused on analyzing data in real-time.
        \item \ul{Similarity}: Both ECA rules and continuous queries can be used to process data in real-time.
    \end{itemize}
\end{ejercicio}

\newpage\subsection{IoT Machine Learning}

The recorded lecture is provided \myhref{https://www.youtube.com/watch?v=S4ycy_yPp2c&feature=youtu.be}{here}.

\begin{ejercicio}
    Why would we want to move machine learning to the edge or even into the smart IoT thing? What are potential advantages? What are challenges and when would you not move it but keep it in the Cloud?\\

    Moving machine learning to the edge or into the smart IoT thing can have several advantages, such as:
    \begin{itemize}
        \item Reduced latency
        \item Reduced bandwidth usage
        \item Improved privacy and security
    \end{itemize}
    However, there are also challenges, such as:
    \begin{itemize}
        \item Limited computational resources
        \item Limited energy resources
        \item Limited storage resources
    \end{itemize}
    We would not move machine learning to the edge or into the smart IoT thing if the computational, energy, or storage requirements of the machine learning model exceed the capabilities of the edge device or smart IoT thing. In such cases, it would be more efficient to keep the machine learning model in the Cloud, where there are more resources available to handle the computational demands of the model.
\end{ejercicio}

\begin{ejercicio}
    Briefly explain how magnitude-based pruning is performed.

    Magnitude-based pruning is a technique used to reduce the size of a neural network by removing weights that have a small magnitude. The idea is that weights with small magnitudes (close to zero) have less impact on the output of the network and can be removed without significantly affecting the performance of the model. As every prunning technique, it will then require a fine-tuning step to recover the performance of the model after pruning.
\end{ejercicio}

\begin{ejercicio}
    Explain the difference between pruning and quantization of an artificial neural network. What is binarization and why has it the potential to be implemented extremely efficiently in hardware?
    
    Pruning is a technique used to reduce the number of calculations in a neural network by removing weights that have a small impact on the output of the network. Quantization, on the other hand, is a technique used to reduce the cost of each computation by reducing the precision of the weights and activations in the neural network. Binarization is a extreme type of quantization where the weights and activations are reduced to binary values (0 and 1). Binarization has the potential to be implemented extremely efficiently in hardware because it can leverage bitwise operations, which are much faster than floating-point operations used in higher-precision models. This can lead to significant speedups and energy savings, especially in edge devices with limited computational resources.
\end{ejercicio}