\section{IoT Internetworking}

\subsection{Motivation}

The recorded lecture is provided \myhref{https://www.youtube.com/watch?v=9YRa5QyNf4w&feature=youtu.be}{here}.

\begin{ejercicio}
    Why do we need new protocols for the IoT instead of using established Internet protocols? Describe one challenge existing protocols face.\\

    New protocols are needed for the IoT because existing Internet protocols, such as TCP/IP, are not optimized for the resource-constrained nature of IoT devices. One challenge existing protocols face is that they may require larger buffer sizes than what is available on IoT devices, which can lead to performance issues and increased latency.
\end{ejercicio}

\begin{ejercicio}
    Which central protocol can we not replace (completely) for the IoT? Why is it harder to replace it than others?\\

    The central protocol that we cannot replace (completely) for the IoT is the Internet Protocol (IP). It is harder to replace IP than other protocols because it is the fundamental protocol that enables communication and data exchange between different devices and applications on the Internet. It is the backbone of the Internet and, technically speaking, being connected to the Internet requires having an IP address and using IP for communication. Therefore, while we can use different protocols for the application layer or the transport layer, we still need to use IP at the network layer in order to enable connectivity and communication with the Internet. In addition, replacing IP would require a complete overhaul of the Internet infrastructure and would require all devices and applications to adopt the new protocol, which is a significant challenge in terms of compatibility and interoperability.
\end{ejercicio}

\begin{ejercicio}
    We introduced two approaches for Smart Things to connect to the Internet. Name these approaches, describe them briefly and give one advantage and one disadvantage of each, compared to the other.\\

    The two approaches for Smart Things to connect to the Internet are:
    \begin{itemize}
        \item \ul{Application layer proxy}: In this approach, there is a proxy (usually implemented in the IoT gateway) that translates between the protocols used by the IoT Things and the protocols used by the Internet. The IoT Thing communicates with the proxy using dedicated IoT protocols that have nothing to do with the Internet protocols, and the proxy stores the data received from the IoT Things and makes it available to the Internet using standard Internet protocols.
        \begin{itemize}
            \item Advantage: It allows for the use of dedicated IoT protocols that are optimized for the resource-constrained nature of IoT devices, which can lead to better performance and energy efficiency.
            \item Disadvantage: It introduces an additional layer of complexity and can lead to increased latency and reduced scalability (due to the dedicated software), as all communication between the IoT Things and the Internet needs to go through the proxy.
        \end{itemize}

        \item \ul{IP Router}: In this approach, the IoT Things directly use IP, but with some optimizations and adaptations to address the challenges and limitations mentioned above. There is no need for specific apps or software to interact with the IoT Things, as they can be accessed using standard Internet protocols and interfaces.
        \begin{itemize}
            \item Advantage: It provides better scalability and interoperability, as the IoT Things can communicate directly with other devices and applications on the Internet without the need for a proxy.
            \item Disadvantage: It may require more complex and resource-intensive protocols and technologies to enable efficient communication and data exchange between the IoT Things and the Internet, which can lead to increased power consumption and reduced performance for resource-constrained IoT devices.
        \end{itemize}
    \end{itemize}
\end{ejercicio}

\newpage\subsection{IEEE 802.15.4}

The recorded lectures are provided \myhref{https://www.youtube.com/watch?v=utk_hszYO_o}{here (Part A)} and \myhref{https://www.youtube.com/watch?v=dOYCUwEpEnw}{here (Part B)}.

\begin{ejercicio}
    What are the differences between the PHY and the MAC layer?\\

    Both are layers in the OSI model, but they serve different purposes. The PHY (Physical) layer is responsible for the physical transmission of data over the communication medium. The MAC (Medium Access Control) layer builds on top of the PHY layer and is responsible for managing access to the communication medium, including addressing, framing, and error detection.
\end{ejercicio}

\begin{ejercicio}
    Discuss pros and cons of using 16 bit addresses for 802.15.4 instead of 64 bit addresses.\\

    Some pros of using 16 bit addresses for 802.15.4 include:
    \begin{itemize}
        \item Reduced overhead: 16 bit addresses require less memory and bandwidth compared to 64 bit addresses.
        \item Faster processing: 16 bit addresses can be processed more quickly than 64 bit addresses, which can lead to improved performance and reduced latency.
    \end{itemize}

    Some cons of using 16 bit addresses for 802.15.4 include:
    \begin{itemize}
        \item Limited address space: 16 bit addresses can only support up to $2^{16} = 65536$ unique addresses, which may not be sufficient for large-scale IoT deployments.
        \item Not globally unique: 16 bit addresses are not globally unique, so they can only be used within a local network, which can limit interoperability and scalability.
        \item Harder to manage: They have to be managed and assigned manually, which can lead to errors and conflicts in larger networks.
        \item If a device moves to another network, it may need to change its address, which can lead to additional overhead and complexity in managing the network.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    Describe the basic idea of CSMA/CA and compare it to frequency hopping. You may use a diagram / drawing to describe your answer.\\

    CSMA/CA (Carrier Sense Multiple Access with Collision Avoidance) is a protocol used in wireless communication to manage access to the communication medium. It works by having devices listen to the channel before transmitting data. If the channel is busy, the device will wait for a random backoff time before trying again. This helps to avoid collisions and improve the efficiency of the network. In contrast, frequency hopping is a technique where devices rapidly switch between different frequency channels to avoid interference and improve security. The hopping pattern is typically determined by a pseudo-random sequence, so no two devices will hop to the same frequency at the same time, which means that they can transmit simultaneously without interference.
\end{ejercicio}

\begin{ejercicio}
    Briefly describe and compare TDMA and FDMA. Which one is better?\\

    TDMA (Time Division Multiple Access) is a protocol that divides the communication medium into time slots, and each device is assigned the whole frequency channel for a specific time slot. This means that, while only one device can transmit at a time, it can use the full bandwidth of the channel during its assigned time slot. FDMA (Frequency Division Multiple Access) is a protocol that divides the communication medium into frequency channels, and each device is assigned a specific frequency channel for transmission. This means that multiple devices can transmit simultaneously, but they have to share the bandwidth of the channel.
    The choice between TDMA and FDMA depends on the specific requirements of the application and the characteristics of the communication medium. 
    \begin{itemize}
        \item Depending on the application.
        
        TDMA may be better for applications that require high data rates but can tolerate some delay, while FDMA may be better for applications that require no delay but can tolerate lower data rates.

        \item Depending on the channel characteristics.
        
        If the bandwidth of the channel is limited, TDMA may be more efficient as it allows devices to use the full bandwidth during their assigned time slots. On the other hand, if the channel has a large bandwidth and can support multiple simultaneous transmissions, FDMA may be more efficient as it allows multiple devices to transmit at the same time without interference.

        \item Depending on the number of devices.
        
        If there are a large number of devices that need to access the communication medium, FDMA may be more efficient as it allows multiple devices to transmit simultaneously, while TDMA may lead to increased latency and reduced performance due to the need for devices to wait for their assigned time slots.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    An 802.15.4 network is deployed in the following three application scenarios. Which CSMA/CA MAC Layer version would you use? Justify your answer (e.g., ``In this case, we should use slotted CSMA/CA, because...'').
    \begin{itemize}
        \item A smart home network, where only 10 temperature sensors are installed. Each sensor reports a measurement once per minute.
        
        Given the low number of devices, the low data rate, and the fact that the devices are not synchronized, we can use unslotted CSMA/CA, as it is simpler and more efficient for this scenario.
        \item A smart clinic network that transmits monitoring data about the patients and alarm notifications.
        
        In this case, we should use slotted CSMA/CA, because it can provide better performance and reduced latency for time-sensitive applications, such as monitoring data. In addition, thanks to the GTS mechanism, it can provide guaranteed time slots for alarm notifications, which can be critical for patient safety.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    When unslotted CSMA/CA is applied, at each trial the back-off window size will be doubled. However, it doesn't always lead to a longer delay, why?\\

    The back-off window size is doubled after each failed transmission attempt, but this does not always lead to a longer delay because the back-off time is randomly selected from the back-off window. The expected back-off time may be doubled, but it is possible that the randomly selected back-off time is shorter than the previous one, which can lead to a shorter delay.
\end{ejercicio}

\newpage\subsection{NB-IoT}

The recorded lectures are provided \myhref{https://www.youtube.com/watch?v=TJFb45bsDVs}{here (Part A)} and \myhref{https://www.youtube.com/watch?v=IGR7OcdVX84}{here (Part B)}.

\begin{ejercicio}
    Explain the basic architecture of a cellular network. How does it cover large areas?

    A cellular network is a wireless communication network that is divided into smaller geographic areas called cells. Each cell is served by a base station, which is responsible for providing wireless coverage and connectivity to the devices within its cell. The base stations are connected to a central network infrastructure, which manages the communication and data exchange between the devices and the Internet. The cellular network covers large areas by using multiple cells that are strategically placed to provide coverage to the desired geographic area. The cells can be of different sizes, depending on the population density and the demand for wireless services in the area. By using multiple cells, the cellular network can provide coverage to a large number of devices while maintaining good performance and reliability.
\end{ejercicio}

\begin{ejercicio}
    Briefly describe an application case in which NB-IoT is better than IEEE 802.15.4. Justify your answer.

    An application case in which NB-IoT is better than IEEE 802.15.4 is whenever you need to cover a large area with a low density of devices, such as in smart agriculture or smart city applications. NB-IoT is designed for wide-area coverage and can support a large number of devices over a wide geographic area, while IEEE 802.15.4 is designed for short-range communication and is better suited for applications that require low power consumption and low data rates. However, IEEE 802.15.4 should not be forgotten, as it works in the unlicensed spectrum and can be used without the need for a cellular network infrastructure, which can be advantageous in certain scenarios, such as in remote areas or in applications that require low-cost deployment.
\end{ejercicio}

\begin{ejercicio}
    Name and briefly explain two of the possible deployment scenarios that specify the frequency range that NB-IoT may be deployed in.

    Two possible scenarios for NB-IoT deployment are:
    \begin{itemize}
        \item \ul{In-band deployment}: In this scenario, NB-IoT is deployed within the existing LTE spectrum, using the same frequency bands as LTE. This allows for easy integration with existing LTE infrastructure and can provide good performance and coverage for NB-IoT devices. However, it may lead to increased interference and reduced performance for both NB-IoT and LTE devices, as they are sharing the same frequency bands.
        \item \ul{Guard-band deployment}: In this scenario, NB-IoT is deployed in the guard bands between LTE frequency bands. This allows for better isolation and reduced interference between NB-IoT and LTE devices, as they are using different frequency bands. However, it may require additional spectrum allocation and may not be as widely supported as in-band deployment, which can limit the availability and coverage of NB-IoT devices.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    How are frequency and time multiplexing used in NB-IoT?

    NB-IoT uses both frequency and time multiplexing to enable efficient communication and data exchange between the devices and the base station.
    \begin{itemize}
        \item \ul{Frequency multiplexing}: NB-IoT uses frequency division multiplexing (FDM) to divide the available frequency spectrum into two different types of channels: uplink channels for transmitting data from the devices to the base station, and downlink channels for transmitting data from the base station to the devices. 
        \item \ul{Time multiplexing}: NB-IoT uses time division multiplexing (TDM) to divide the physical channel into time slots, and each time slot is assigned to a specific logical channel. This allows multiple devices to share the same frequency channel by transmitting and receiving data in different time slots.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    Explain the procedure that a newly powered on user equipment (UE) follows to join a cell.

    When a newly powered on user equipment (UE) wants to join a cell, it follows the following procedure:
\begin{itemize}
    \item The UE listens in a broadcast channel for a signal from base stations, which contains information about the available cell and its characteristics.
    \item The UE checks if the cell is joinable, as some cells may be reserved for specific types of devices or applications.
    \item If the cell is joinable, the UE checks if it has enough signal strength to connect to the cell.
    \item Among the cells that are joinable and have enough signal strength, the UE selects the one with the best signal quality (e.g., the one with the highest signal-to-noise ratio) and sends a connection request to the base station of that cell.
\end{itemize}
\end{ejercicio}

\begin{ejercicio}
    What is a tracking area in NB-IoT and how is it used?

    A tracking area in NB-IoT is a group of cells that share the same tracking area identifier (TAI). The MME (Mobility Management Entity) keeps track of the tracking area that is serving the UE, so when there is data waiting for the UE, only the cells in that tracking area are paged. When the UE needs to inform the MME of the cell that is serving it, it sends a Tracking Area Update (TAU) message to the MME, which updates in its internal records the tracking area that is serving the UE. This allows for a trade-off between the overhead of paging and the overhead of tracking, as it reduces the overhead of paging while still allowing for efficient tracking of the UE's location within the network.
\end{ejercicio}

\begin{ejercicio}
    How can a UE use Extended Discontinuous Reception to save energy?

    Extended Discontinuous Reception (eDRX) is a mode that allows the UE to save energy by sleeping not just between the start of each hyperframe (when paging is performed), but for a configurable number of hyperframes. This means that the UE can configure the number of hyperframes that it will sleep for, so it will only listen for paging messages at the start of each hyperframe after sleeping for that number of hyperframes. This allows the UE to save energy by sleeping for longer periods of time, while still being able to receive paging messages from the base station when needed. During all that time, the SGW retains the incoming data for the UE in buffers.
\end{ejercicio}


\newpage\subsection{6LoWPAN}

The recorded lectures are provided \myhref{https://www.youtube.com/watch?v=krkHR0lR2Xg}{here (Part A)} and \myhref{https://www.youtube.com/watch?v=EcPQGWNfqns}{here (Part B)}.

\begin{ejercicio}
    Explain why we need 6LoWPAN to support IPv6 over IEEE 802.15.4. Also explain briefly how 6LoWPAN can make IPv6 over IEEE 802.15.4 more efficient.\\

    We need 6LoWPAN to support IPv6 over IEEE 802.15.4 strictly just because of one reason: IP requires a minimum MTU of 1280 bytes, while IEEE 802.15.4 has a maximum frame size of 127 bytes. This means that we cannot directly use IPv6 over IEEE 802.15.4 without some form of adaptation or optimization, as the IPv6 packets would be too large to fit within the frame size of IEEE 802.15.4. 6LoWPAN provides fragmentation in order to transparently split IPv6 packets into smaller fragments that can fit within the frame size of IEEE 802.15.4.

    However, in order to make IPv6 over IEEE 802.15.4 more efficient, 6LoWPAN also provides two main optimizations:
    \begin{itemize}
        \item \ul{Header compression}: 6LoWPAN provides header compression techniques that can significantly reduce the size of the IPv6 header, which can increase the payload rate and reduce the overhead of transmitting IPv6 packets over IEEE 802.15.4.
        \item \ul{Stateless Address Autoconfiguration}: 6LoWPAN provides a mechanism for stateless address autoconfiguration, which allows devices to automatically generate their own IPv6 addresses based on their MAC addresses and the LL prefix. This can simplify the configuration and management of IPv6 addresses for devices in an IEEE 802.15.4 network, and can also reduce the overhead of address management and configuration.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    What is an Encapsulation Header Stack in 6LoWPAN?

    An Encapsulation Header Stack in 6LoWPAN is a stack of headers that are added to the IPv6 packet in order to provide additional information and functionality for the transmission of the packet over IEEE 802.15.4. The Encapsulation Header Stack can include headers for fragmentation and header compression, as well as other headers that may be needed for specific applications or use cases. The Encapsulation Header Stack is added to the IPv6 packet before it is transmitted over IEEE 802.15.4, and it is removed at the receiving end before the IPv6 packet is processed by the upper layers of the protocol stack.
\end{ejercicio}

\begin{ejercicio}
    Explain the concept of ``Stateless Address Autoconfiguration'' for 6LoWPAN, without going into detail about how an address is created.

    Stateless Address Autoconfiguration for 6LoWPAN is a mechanism that allows devices to automatically generate their own IPv6 addresses without the need for a central address management system, as DHCP does. The addresses generated by this mechanism are however link-local addresses, which means that they can only be used for communication within the same link (e.g., within the same IEEE 802.15.4 network), and they cannot be used for communication with devices outside of the link (e.g., with devices on the Internet). This mechanism is particularly useful for resource-constrained devices in an IEEE 802.15.4 network, as it allows them to generate their own addresses without the need for additional configuration or management overhead.
\end{ejercicio}

\begin{ejercicio}
    Explain briefly how fragmentation in 6LowPAN relates to fragmentation in IP. Does it, e.g., replace it?

    Given that the maximum frame size of IEEE 802.15.4 is much smaller than the minimum MTU required by IP, fragmentation in 6LoWPAN is necessary to enable the transmission of IPv6 packets over IEEE 802.15.4. However, fragmentation in 6LoWPAN does not replace fragmentation in IP, as the longest possible IPv6 packet that can be fragmented by 6LoWPAN is $2^{11}-1 = 2047$ bytes, which is still small for the maximum size of an IPv6 packet (which can be up to $2^{16}-1 = 65535$ bytes). Therefore, if an IPv6 packet is larger than 2047 bytes, it would still need to be fragmented at the IP layer in order to be transmitted over IEEE 802.15.4, and the resulting fragments would then be further fragmented by 6LoWPAN in order to fit within the frame size of IEEE 802.15.4.

    % // TODO: Ver TODOs
\end{ejercicio}

\begin{ejercicio}
    A single IPv6 packet with a size of 555 Byte (including the IPv6 header and its payload) is sent to an IoT device using 802.15.4 and 6LoWPAN. The IPv6 header is uncompressed, and we use 6LoWPAN fragmentation. What would be the value of the datagram offset field in the 5th fragment? Justify your answer.
    \begin{observacion}
        We assume each fragment has 72 bytes available for its payload.
    \end{observacion}
    \begin{enumerate}
        \item For the first fragment, given that $72=9\cdot 8$ is divisible by 8, we can use all the 72 bytes for the payload, so the datagram contains bytes 0-71 of the original IPv6 packet. It has no offset, as it is the first fragment.
        \item For the second fragment, we can also use all the 72 bytes for the payload, so the datagram contains bytes 72-143 of the original IPv6 packet. The datagram offset is 9, as $\nicefrac{72}{8}=9$.
        \item For the third fragment, we can also use all the 72 bytes for the payload, so the datagram contains bytes 144-215 of the original IPv6 packet. The datagram offset is 18, as $\nicefrac{144}{8}=18$.
        \item For the fourth fragment, we can also use all the 72 bytes for the payload, so the datagram contains bytes 216-287 of the original IPv6 packet. The datagram offset is 27, as $\nicefrac{216}{8}=27$.
        \item For the fifth fragment, we can also use all the 72 bytes for the payload, so the datagram contains bytes 288-359 of the original IPv6 packet. The datagram offset is 36, as $\nicefrac{288}{8}=36$.
    \end{enumerate}
\end{ejercicio}

\begin{ejercicio}
    Explain when using HC1 and HC2 for header compression is especially useful and when it is not very useful.

    Using HC1 and HC2 for header compression is especially useful when the IPv6 packets being transmitted over IEEE 802.15.4 have LL auto-configured source and destination addresses, as these addresses can be compressed to two bits each using HC1, which can significantly reduce the size of the IPv6 header and increase the payload rate. In addition, they are useful when the next header is either UDP, ICMP, or TCP, as these can also be compressed using HC2, which can further reduce the size of the header and increase the payload rate. In other cases, they may not be very useful.
\end{ejercicio}







\newpage\subsection{MQTT}

The recorded lecture is provided \myhref{https://www.youtube.com/watch?v=4_PkeL20LB8&feature=youtu.be}{here}.

\begin{ejercicio}
    Given the following MQTT configuration with five clients $C1, C2, C3, C4, C5$ and a broker $B$, which clients receive published messages in the following scenarios?
    \begin{enumerate}
        \item $C1$ publishes a message for topic \texttt{home/groundfloor/temp} with the retain flag set to ``1''. $C2$ publishes a message for topic \texttt{home/secondfloor/temp} with the retain flag set to ``0''. $C3$ publishes a message for topic \texttt{home/thirdfloor/livingroom/temp} with the retain flag set to ``1''. After all publications are done, $C4$ subscribes to topic \texttt{home/secondfloor/temp} and $C5$ subscribes to topic \texttt{home}.
        \item $C1$ subscribes to topic \texttt{home/thirdfloor/+/temp}. After this, $C2$ publishes a message for topic \texttt{home/groundfloor/temp} with the retain flag set to ``1''. $C3$ publishes a message for topic \texttt{home/secondfloor/temp} with the retain flag set to ``0''. $C4$ publishes a message for topic \texttt{home/thirdfloor/livingroom/temp} with the retain flag set to ``0''. After all publications done, $C5$ subscribes to topic \texttt{home/thirdfloor/livingroom/temp}.
        \item $C1$ subscribes to topic \texttt{home/groundfloor/temp}. $C2$ subscribes to topic \texttt{home/secondfloor/temp}. After the subscriptions are done, $C3$ publishes a message for topic \texttt{home/+/temp}.
        \item $C1$ subscribes to topic \texttt{home/\#} with QoS level 0. $C2$ subscribes to topic \texttt{home/+/temp} with QoS level 0. After the subscriptions are done, $C3$ publishes a message for topic \texttt{home/thirdfloor/temp} with QoS level 2.
    \end{enumerate}
\end{ejercicio}

\begin{ejercicio}
    Client $C2$ subscribes to the topic \texttt{a/b/c} with QoS level 0. Afterwards, client $C1$ publishes a message to the topic \texttt{a/b/c} with QoS level 2. Looking at the message headers: what is the QoS level between $C1$ and the broker? What is the QoS level between the broker and $C2$?
\end{ejercicio}

\begin{ejercicio}
    Explain how to use ``Shared Subscriptions'' in MQTT.
\end{ejercicio}