\chapter{Search Algorithms}

In our current world, plenty of algorithms are used to search for information. From the most basic ones, such as sorting algorithms, to the most complex ones, such as search engines. In this chapter, we will explore some of the most common search algorithms and their applications.

\section{Graph Theory}

Search algorithms are often used in graphs, as they provide a way to represent complex relationships between data. A graph $G=(V,E)$ is defined as a pair of sets, where $V$ is called the set of vertices and $E$ is called the set of edges. The vertices represent the data points, while the edges represent the relationships between them. There are different types of graphs, such as directed and undirected graphs, weighted and unweighted graphs, or cyclic and acyclic graphs. Each type of graph has its own properties and algorithms that can be applied to it. An example graph is shown in Figure~\ref{fig:graph}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[circle, draw] (A) at (0,0) {A};
        \node[circle, draw] (B) at (0, -2) {B};
        \node[circle, draw] (C) at (2, -2) {C};
        \node[circle, draw] (D) at (2, -0.7) {D};
        \node[circle, draw] (E) at (4, 0) {E};
        \draw[Stealth-Stealth] (A) to (B);
        \draw[Stealth-Stealth] (A) to (E);
        \draw[Stealth-Stealth] (A) to (D);
        \draw[Stealth-Stealth] (B) to (D);
        \draw[Stealth-Stealth] (D) to (E);
        \draw[Stealth-Stealth] (C) to (E);
        \draw[-Stealth] (B) to (C);
    \end{tikzpicture}
    \caption{Example of a directed graph.}
    \label{fig:graph}
\end{figure}

A graph is usually represented as an \emph{adjacency matrix}. Where the rows represent the origin vertices, the columns represent the destination vertices\footnote{This could also be the opposite: the origin vertices could be the colums and the destination vertices, the rows.}, and the values $a_{ij}$ represent:
\begin{itemize}
    \item If the graph is unweighted, $a_{ij}$ represents the number of edges from vertex $i$ to vertex $j$.
    \item If the graph is weighted, $a_{ij}$ represents the weight of the edge from vertex $i$ to vertex $j$. If there is no edge, $a_{ij}$ is usually set to infinity or an invalid value.
\end{itemize}
If there are a lot of vertices that are not connected to each other, the adjacency matrix can be very sparse, meaning that it contains a lot of zeros. In this case, it is more efficient to use an \emph{adjacency list} to represent the graph, where each vertex is represented as a list of its neighboring vertices. This can save a lot of space, especially for large graphs with few edges. For example, the adjacency matrix for the graph in Figure~\ref{fig:graph} is shown in Equation~\ref{eq:adjacency}, while the adjacency list is Listing~\ref{lst:adjacency_list}.
\begin{equation}
    A = \begin{bmatrix}
        0 & 1 & 0 & 1 & 1 \\
        1 & 0 & 1 & 1 & 0 \\
        0 & 0 & 0 & 0 & 1 \\
        1 & 1 & 0 & 0 & 1 \\
        1 & 0 & 1 & 1 & 0 \\
    \end{bmatrix}
    \label{eq:adjacency}
\end{equation}
\begin{listing}
    \begin{minted}{python}
        adjacency_list = {
            'A': ['B', 'D', 'E'],
            'B': ['A', 'C', 'D'],
            'C': ['B', 'E'],
            'D': ['A', 'B', 'E'],
            'E': ['A', 'C', 'D']
        }
    \end{minted}
    \caption{Adjacency list for the graph in Figure~\ref{fig:graph}.}
    \label{lst:adjacency_list}
\end{listing}

This matrix is also called ``next-neighbor'' matrix, as the neighboring vertices of a vertex $i$ can be found by looking at the $i$-th row of the matrix. The ``next-next-neighbor'' matrix, on the other hand, is obtained by multiplying the adjacency matrix by itself. This matrix can be used to find the vertices that are two steps away from a given vertex.

Lastly, the \emph{transition matrix} is obtained by normalizing the adjacency matrix. This is done by dividing each entry $a_{ij}$ by the sum of the entries in the $i$-th row. The transition matrix can be used to find the probability of transitioning from one vertex to another in a random decision process.


\subsection{PageRank}

PageRank is an algorithm used to determine the importance of nodes in a graph. It is well-known for being used by Google Search to rank web pages in their search engine results. It works by counting the number and quality of links to a page to determine a rough estimate of how important the website is. The underlying assumption is that more important websites are likely to receive more links from other websites. Everything is represented as a directed graph, where the vertices represent web pages and the edges represent hyperlinks between them. There are certain parameters that can be adjusted in the PageRank algorithm, such as:
\begin{itemize}
    \item The teleportation probability vector. It's $n$-th entry represents the probability of jumping to the $n$-th page when a random surfer decides to teleport (going to a random page instead of following a link). It is usually set to a uniform distribution, meaning that the random surfer has an equal probability of jumping to any page.
    \item The damping factor $\alpha$, which represents the probability that a random surfer will continue following links rather than teleporting.
    \item The convergence threshold, which determines when the algorithm should stop iterating and return the PageRank values.
\end{itemize}

The algorithm returns the PageRank vector, whose $n$-th entry represents the PageRank of the $n$-th page. The PageRank values decide in which order the pages are shown in the search results, with higher PageRank values indicating more important pages. The PageRank algorithm works as follows:
\begin{enumerate}
    \item Initialize the PageRank vector with the teleportation probability vector.
    \item Update the old PageRank vector (PR) by adding the transition matrix multiplied by the old PageRank vector (weighted by the damping factor $\alpha$) and the teleportation probability vector (weighted by $1-\alpha$):
    \begin{equation*}
        PR = \alpha \cdot M \cdot PR + (1 - \alpha) \cdot v
    \end{equation*}
    Where $M$ is the transition matrix and $v$ is the teleportation probability vector.
    \item Repeat the process until the change in the PageRank vector is less than the convergence threshold (or a maximum number of iterations is reached).
\end{enumerate}

Once the algorithm converges, the PageRank values can be used to rank the web pages in search results. The higher the PageRank value, the more important the page is considered to be, and therefore, it will be ranked higher in the search results.

\section{Sorting Algorithms}

Sorting algorithms are a fundamental part of computer science, as they allow us to organize data in a specific order. In addition, they are used in many search algorithms, as they can help to reduce the search space and improve the efficiency of the search.

\subsection{Insertion Sort}

This algorithm works by dividing the array into a sorted and an unsorted part. The sorted part is initially empty, and the unsorted part contains all the elements.
\begin{enumerate}
    \item Start with the first element of the unsorted part and compare it with the last element of the sorted part.
    \item IMove all the elements of the sorted part that are greater than the current element to the right.
    \item Insert the current element into its correct position in the sorted part.
    \item Repeat the process for all the elements in the unsorted part until it is empty.
\end{enumerate}

It has a time complexity of $O(n^2)$.

\subsection{Bubble Sort}

This algorithm works by repeatedly swapping adjacent elements if they are in the wrong order. The process is repeated until the array is sorted.
\begin{enumerate}
    \item Compare each pair of adjacent elements in the array.
    \item If the elements are in the wrong order, swap them.
    \item Repeat the process for all the elements in the array until it is sorted.
\end{enumerate}
It has a time complexity of $O(n^2)$.

\subsection{Shaker Sort}

This algorithm is a variation of Bubble Sort that sorts in both directions on each pass through the list. It works by comparing adjacent elements and swapping them if they are in the wrong order, similar to Bubble Sort. However, it also compares elements in the opposite direction, which can help to reduce the number of passes needed to sort the array.
\begin{enumerate}
    \item Start at the beginning of the array and compare each pair of adjacent elements.
    \item If the elements are in the wrong order, swap them.
    \item After reaching the end of the array, reverse the direction and compare each pair of adjacent elements again.
    \item Repeat the process until the array is sorted.
\end{enumerate}
It has a time complexity of $O(n^2)$, but with lower constant factors than Bubble Sort.

\subsection{Quick Sort}

This divide-and-conquer algorithm works by selecting a pivot element from the array and partitioning the other elements into two sub-arrays, according to whether they are less than or greater than the pivot. The sub-arrays are then sorted recursively. This process continues until the base case of an empty array or an array with a single element is reached.
\begin{enumerate}
    \item Choose a pivot element from the array.
    \item Partition the array into two sub-arrays: one with elements less than the pivot and one with elements greater than the pivot.
    \item At this point, the pivot is in its final position.
    \item Recursively apply the above steps to the sub-arrays until the base case is reached.
\end{enumerate}

The efficiency of Quick Sort depends on the choice of the pivot. The complexity depends on the partitioning of the array. If the pivot divides the array into two equal halves, the time complexity is $O(n \log n)$. However, if the pivot is consistently the smallest or largest element, the time complexity degrades to $O(n^2)$.

\subsection{Topological Sorting: Kahn's Algorithm}

This algorithm is used to sort the vertices of a directed acyclic graph (DAG) in a linear order. If there is a directed edge from vertex $u$ to vertex $v$, then $u$ comes before $v$ in the ordering. It is used in several applications, such as scheduling tasks, resolving dependencies, and determining the order of compilation in programming languages. The algorithm works as follows:
\begin{enumerate}
    \item The result is going to be stored in a list called $L$.
    \item Find all vertices with no incoming edges and add them to a set $S$.
    \item While $S$ is not empty:
    \begin{enumerate}
        \item Remove a vertex $n$ from $S$ and add it to $L$.
        \item For each vertex $m$ with an edge from $n$ to $m$:
        \begin{enumerate}
            \item Remove the edge from $n$ to $m$.
            \item If $m$ has no other incoming edges, add $m$ to $S$.
        \end{enumerate}
    \end{enumerate}
    \item If there are still edges in the graph, it means that there is a cycle, and therefore, a topological sort is not possible. Otherwise, the list $L$ contains the vertices in topologically sorted order.
\end{enumerate}

\section{Uninformed Search}

Uninformed search algorithms are those that do not have any additional information about the search space. They explore the search space in a systematic way, without any guidance or heuristics.

\subsection{Breadth-First Search}

This algorithm explores the search space level by level, starting from the root node and expanding all the nodes at the present depth before moving on to the nodes at the next depth level. It uses a queue (FIFO) to keep track of the nodes to be explored. It is guaranteed to find the shortest path, but it can be memory intensive, as it needs to store all the nodes at the current depth level.

\subsection{Depth-First Search}

This algorithm explores the search space by going as deep as possible along each branch before backtracking. It uses a stack (LIFO) to keep track of the nodes to be explored. It is not guaranteed to find the shortest path, but it can be more memory efficient than breadth-first search.

\subsection{Depth-Limited Search}

This algorithm is a variation of depth-first search that limits the depth of the search. It is used to avoid infinite loops in cases where the search space is infinite or has cycles. It uses a stack (LIFO) to keep track of the nodes to be explored, and it keeps track of the current depth of the search. If the depth limit is reached, it backtracks and explores other branches. It is not guaranteed to find the searched node, but it can be more memory efficient than depth-first search. The choice of the depth limit is crucial, as it can affect the completeness and optimality of the search.

\subsection{Iterative Deepening Search}

This algorithm combines the benefits of both breadth-first and depth-first search. It performs a depth-limited search with increasing depth limits until a solution is found.
\begin{enumerate}
    \item Start with a depth limit of 0.
    \item Perform a depth-limited search with the current depth limit.
    \item If a solution is found, return it. Otherwise, increase the depth limit and repeat the process.
\end{enumerate}

A stack (LIFO) is used to keep track of the nodes to be explored in the depth-limited search. It is guaranteed to find the shortest path, and it can be more memory efficient than breadth-first search.

\subsection{Bidirectional Search}

This algorithm is used to find the shortest path between two nodes in a graph. It works by simultaneously searching (typically BFS) from both the start node and the goal node until the two searches meet in the middle. It uses two data structures to keep track of the nodes to be explored from both directions. It is guaranteed to find the shortest path, and it can be more efficient than unidirectional search, especially in cases where the search space is large.

\subsection{Uniform Cost Search}

This algorithm is used in weighted graphs (where the edges have positive costs) to find the least costly path from the root node to a goal node. It is similar to breadth-first search, but it takes into account the cost of the path.
It uses a priority queue to keep track of the nodes to be explored, where the priority is determined by the cost of the path from the root node to the current node. It is guaranteed to find the least costly path, but it can be memory intensive, as it needs to store all the nodes in the search space.
\begin{enumerate}
    \item Start with the root node and add it to the priority queue with a cost of 0.
    \item While the priority queue is not empty:
    \begin{enumerate}
        \item Remove the node with the lowest cost from the priority queue.
        \item If the node is a goal node, return the path to it.
        \item Otherwise, expand the node and add its children to the priority queue with their respective cumulative costs.
    \end{enumerate}
\end{enumerate}


\subsubsection{Local Uniform Cost Search}

This algorithm is a variation of uniform cost search that is used to find the least costly path from a given node to a goal node, but just considering the local neighborhood of the given node. For each selected node, the next selected node is the one with the lowest cost among its neighbors. It is not guaranteed to find the least costly path, but it can be more efficient than uniform cost search in cases where the search space is large and the goal node is close to the given node.
\begin{enumerate}
    \item Start with the given node.
    \item While the goal node has not been reached, or a maximum number of iterations has not been reached:
    \begin{enumerate}
        \item Evaluate the neighbors of the current node and select the one with the lowest cost.
        \item Move to the selected neighbor and repeat the process.
    \end{enumerate} 
\end{enumerate}

\subsection{Dijkstra's Algorithm}

This algorithm is a specific implementation of uniform cost search that is used to find the shortest path from a source node to all other nodes in a weighted graph. It works by maintaining a priority queue of nodes to be explored, where the priority is determined by the cumulative cost from the source node to the current node. The algorithm iteratively expands the node with the lowest cumulative cost and updates the costs of its neighbors until all nodes have been explored. It is guaranteed to find the shortest path from the source node to all other nodes, but it can be memory intensive, as it needs to store all the nodes in the search space.
\begin{enumerate}
    \item Start with the source node and add it to the priority queue with a cost of 0. Every other node is added to the priority queue with a cost of infinity.
    \item While the priority queue is not empty (it is not empty until all nodes have been explored):
    \begin{enumerate}
        \item Remove the node with the lowest cumulative cost from the priority queue.
        \item For each neighbor of the current node:
        \begin{enumerate}
            \item Calculate the cumulative cost to reach the neighbor through the current node.
            \item If this cumulative cost is less than the previously known cost for that neighbor, update the cost and modify it in the priority queue.
        \end{enumerate}
    \end{enumerate}
\end{enumerate}

It should be noted that Dijkstra's algorithm is the same as UCS, but it simply does not stop when it finds a goal node, but rather continues to explore until all nodes have been explored. This is because Dijkstra's algorithm is designed to find the shortest path from a source node to all other nodes in the graph, while UCS is designed to find the least costly path from a root node to a specific goal node. Therefore, Dijkstra's algorithm is more general than UCS, but also more computationally intensive.


\section{Informed Search}

Informed search algorithms are those that have additional information about the search space. They use this information to guide the search and improve its efficiency. The additional information is usually in the form of a heuristic function, which estimates the cost of reaching the goal from a given node.

\subsection{Greedy Best First Search}

This algorithm uses a priority queue to keep track of the nodes to be explored, where the priority is determined by the heuristic function. It always expands the node that is estimated to be closest to the goal.
\begin{enumerate}
    \item Start with the root node and add it to the priority queue with a priority determined by the heuristic function.
    \item While the priority queue is not empty:
    \begin{enumerate}
        \item Remove the node with the lowest heuristic value from the priority queue.
        \item If the node is a goal node, return the path to it.
        \item Otherwise, expand the node and add its children to the priority queue with their respective heuristic values.
    \end{enumerate}
\end{enumerate}

The efficiency of Greedy Best First Search depends on the quality of the heuristic function. The main drawback of this algorithm is that it is not guaranteed to find the optimal solution, as it does not consider the actual cost of the path, but only its estimated cost to the goal.

\subsection{A* Search}

This algorithm is a combination of uniform cost search and greedy best first search. It uses a priority queue to keep track of the nodes to be explored, where the priority is determined by:
\begin{equation*}
    f(n) = g(n) + h(n)
\end{equation*}
Where $g(n)$ is the actual cost from the root node to the current node $n$, and $h(n)$ is the heuristic estimate of the cost from $n$ to the goal. Two important properties of the heuristic function are:
\begin{itemize}
    \item Admissible: It never overestimates the cost to reach the goal. It can be expressed as:
    \begin{equation*}        h(n) \leq h^*(n)\qquad \forall n\in V
    \end{equation*}
    where $h^*(n)$ is the true cost from node $n$ to the goal. If the heuristic is admissible, A* Search is guaranteed to find the optimal solution.
    \item Consistent (or monotonic): For every node $n$ and every successor $n'$ of $n$, the estimated cost of reaching the goal from $n$ is no greater than the cost of getting from $n$ to $n'$ plus the estimated cost of reaching the goal from $n'$. Formally, this can be expressed as:
    \begin{equation*}
        h(n) \leq c(n, n') + h(n')
    \end{equation*}
    where $c(n, n')$ is the actual cost of moving from node $n$ to node $n'$. If the heuristic is consistent, it is also admissible, and A* Search is guaranteed to find the optimal solution. In addition, if the heuristic is consistent, A* Search is more efficient, as it does not need to revisit nodes that have already been explored.
\end{itemize}

Therefore, the A* Search algorithm is:
\begin{enumerate}
    \item Start with the root node and add it to the priority queue with a priority determined by $f(n) = g(n) + h(n)$. All other nodes are added to the priority queue with a priority of infinity.
    \item While the priority queue is not empty:
    \begin{enumerate}
        \item Remove the node with the lowest $f(n)$ value from the priority queue.
        \item If the node is a goal node, return the path to it.
        \item Otherwise, expand the node. For each child of the current node:
        \begin{enumerate}
            \item Calculate $g(n')$ as $g(n) + c(n, n')$, where $c(n, n')$ is the actual cost of moving from node $n$ to node $n'$. If that new $g(n')$ value is less than the previously known $g(n')$ value for that child, update $g(n')$ and calculate $f(n') = g(n') + h(n')$.
        \end{enumerate}
    \end{enumerate}
\end{enumerate}


\section{Local Search}

Local search algorithms are those that operate on a single solution and try to find a better solution by making small changes to it. They are often used in optimization problems. Therefore, the aim is finding a local optimum, not necessarily a global one.

\subsection{Hill Climbing}

This algorithm starts with an arbitrary solution and iteratively makes small changes to it, accepting the change if it results in a better solution. The process continues until no further improvements can be made (or a maximum number of iterations is reached).
\begin{enumerate}
    \item Start with an arbitrary solution.
    \item Generate the neighbors of the current solution by making small changes to it.
    \item Evaluate the neighbors and select the one with the best value.
    \item If the best neighbor is better than the current solution, move to that neighbor and repeat the process. Otherwise (if no neighbor is better), return the current solution as a local optimum.
\end{enumerate}

\subsection{Simulated Annealing}

This algorithm is a variation of hill climbing that allows for occasional moves to worse solutions in order to escape local optima. It is inspired by the annealing process in metallurgy, where a material is heated and then slowly cooled to remove defects and improve its properties. The algorithm works as follows:
\begin{enumerate}
    \item Start with an arbitrary solution, an initial temperature $T\geq 0$, and a cooling rate $\alpha$ (where $0 < \alpha < 1$).
    \item Generate a neighbor of the current solution by making a small change to it.
    \item Evaluate the neighbor and calculate the change in the objective function $\Delta E$ (difference between the neighbor's value and the current solution's value).
    \item If $\Delta E < 0$ (the neighbor is better), move to that neighbor. Otherwise, move to the neighbor with a probability of $e^{-\Delta E / T}$. Initially, with a high temperature, the algorithm is more likely to accept worse solutions, allowing it to explore the search space more broadly. As the temperature decreases, the algorithm becomes less likely to accept worse solutions, focusing more on local improvements.
    \item Update the temperature by multiplying it by the cooling rate: $T = \alpha T$.
    \item Repeat the process until a stopping criterion is met (e.g., a maximum number of iterations, a minimum temperature, or a satisfactory solution).
\end{enumerate}

This algorithm doesn't necessarily find the optimal solution, but it can avoid being stuck in local optima, especially if the cooling schedule is well-designed.

% \section{Modern utilities}
% // TODO: Desarrollar?

% \subsection{Information Retrieval}

% \subsection{Search Engines}

% \subsection{PageRank}