\chapter{Regression}

During this chapter, we will cover regression, which is a fundamental technique in machine learning used to model and analyze the relationships between variables. We will explore various methods for regression, including linear regression, regularized regression techniques, and outlier detection methods. Additionally, we will discuss classification methods that can be adapted for regression tasks.

\section{Distances}

Before diving into regression, it's essential to understand the concept of distances, as they play a crucial role in many regression algorithms. Different types of distances can be used to measure the similarity or dissimilarity between data points, which can affect the performance of regression models. Four conditions for a function to be considered a distance are:
\begin{enumerate}
    \item \ul{Non-negativity}: $d(x, y) \geq 0$ for all $x, y$.
    \item \ul{Identity of indiscernibles}: $d(x, y) = 0$ if and only if $x = y$.
    \item \ul{Symmetry}: $d(x, y) = d(y, x)$ for all $x, y$.
    \item \ul{Triangle inequality}: $d(x, z) \leq d(x, y) + d(y, z)$ for all $x, y, z$.
\end{enumerate}

Some common types of distances include:
\begin{itemize}
    \item \ul{Euclidean Distance ($L^2$)}: The straight-line distance between two points in a multi-dimensional space, based on the Pythagorean theorem.
    \begin{equation*}
        d(x, y) = \sqrt{\sum_{i=1}^{n} (x_i - y_i)^2}
    \end{equation*}

    \item \ul{Minkowski Distance ($L^p$)}: A generalization of the Euclidean distance that includes other distance metrics as special cases.
    \begin{equation*}
        d(x, y) = \left( \sum_{i=1}^{n} |x_i - y_i|^p \right)^{1/p}
    \end{equation*}

    \item \ul{Manhattan Distance ($L^1$)}: The sum of the absolute differences between the coordinates of two points, often used in grid-like path scenarios.
    \begin{equation*}
        d(x, y) = \sum_{i=1}^{n} |x_i - y_i|
    \end{equation*}

    \item \ul{Max-Distance}: This distance measures the maximum difference between the coordinates of two points, also known as Chebyshev distance.
    \begin{equation*}
        d(x, y) = \max_{i} |x_i - y_i|
    \end{equation*}

    \item \ul{Min-Distance}: It measures the minimum difference between the coordinates of two points.
    \begin{equation*}
        d(x, y) = \min_{i} |x_i - y_i|
    \end{equation*}

    It should be noted that, strictly speaking, the Min-Distance does not satisfy all the conditions to be considered a true distance. However, it is still a useful metric in certain contexts.

    \item \ul{Cosine Similarity}: Measures the cosine of the angle between two vectors, which is a measure of their orientation rather than their magnitude. It is commonly used in text analysis and high-dimensional spaces.
    \begin{equation*}
        d(x, y) = \cos(\theta) = \frac{x \cdot y}{\|x\| \|y\|}
    \end{equation*}
\end{itemize}

Some specific distances for particular types of data include:
\begin{itemize}
    \item \ul{Hamming Distance}: Used for discrete objects or strings, it counts the number of positions at which the corresponding symbols are different.
    \begin{equation*}
        d(x, y) = \sum_{i=1}^{n} \delta(x_i, y_i)
    \end{equation*}
    where $\delta(x_i, y_i) = 1$ if $x_i \neq y_i$ and $0$ otherwise. For instance, the Hamming distance between ``kitten'' and ``sitting'' is 3.

    \item \ul{Mahalanobis Distance}: A distance measure that accounts for the correlations between variables and the scale of the data, making it useful for multivariate data.
    \begin{equation*}
        d(x, y) = \sqrt{(x - y)^T \Sigma^{-1} (x - y)}
    \end{equation*}
    where $\Sigma$ is the covariance matrix of the data:
    \begin{equation*}
        \Sigma = E[(X - \mu)(X - \mu)^T]
    \end{equation*}

    It takes into account the variance of each variable and the covariance between variables, making it particularly effective for identifying outliers in multivariate data.
\end{itemize}

\subsection{Similarity Measures}

In addition to distances, similarity measures are also crucial in regression and other machine learning tasks. They quantify how alike a dataset is, which can be particularly useful in clustering and classification tasks. Some common similarity measures include:
\begin{itemize}
    \item \ul{Entropy}: A measure of uncertainty or randomness in a dataset, often used in decision tree algorithms to determine the best splits.
    \begin{equation*}
        S(X) = -\sum_{x\in X} P(x) \log P(x)
    \end{equation*}
    where $P(x)$ is the probability of occurrence of the value $x$ in the dataset. Depending on the base of the logarithm used, it is measured in bits (base 2), nats (base $e$), or bans (base 10).

    \item \ul{Negentropy}: A measure of the degree of order or structure in a dataset, defined as the difference between the maximum possible entropy and the actual entropy of the dataset.
    \begin{equation*}
        N(X) = S_{\text{max}} - S(X)
    \end{equation*}
    where $S_{\text{max}}$ is the maximum entropy of the dataset, which occurs when all outcomes are equally likely.

    \item \ul{Kullback-Leibler Divergence (Relative Entropy)}: A measure of how one probability distribution $Q(X)$ diverges from a reference probability distribution $P(X)$.
    \begin{equation*}
        D_{KL}(P \| Q) = \sum_{x\in X} P(x) \log \frac{P(x)}{Q(x)}
    \end{equation*}
    It quantifies the amount of information lost when $Q$ is used to approximate $P$.

    \item \ul{Cross-Entropy}: A measure of the entropy of an approximation $Q(X)$ relative to the true distribution $P(X)$, often used as a loss function in classification problems.
    \begin{equation*}
        H(P, Q) = -\sum_{x\in X} P(x) \log Q(x)
    \end{equation*}
    It measures the average number of bits needed to identify an event from a set of possibilities, given a predicted distribution $Q$ instead of the true distribution $P$. It should be noted that:
    \begin{equation*}
        H(P, Q) = S(P) + D_{KL}(P \| Q)
    \end{equation*}
    
    \item \ul{Mutual Information}: A measure of the mutual dependence between two variables, quantifying the amount of information obtained about one variable through the other.
    \begin{equation*}
        MI(X; Y) = \sum_{x \in X} \sum_{y\in Y} P(x, y) \log\left(\frac{P(x, y)}{P(x) P(y)}\right)
    \end{equation*}

    If $MI(X; Y) = 0$, it indicates that the variables are independent, while a higher value indicates a stronger dependency between the variables, and therefore more shared information.
\end{itemize}


\section{Dimensionality Reduction}

Incoming data often has a high number of features, mathematically represented as a high-dimensional space. However, many of these features may be redundant or irrelevant, leading to the curse of dimensionality. Dimensionality reduction techniques aim to reduce the number of features while preserving the essential structure and information in the data. This can improve model performance, reduce overfitting, and enhance interpretability. There are two main types of dimensionality reduction techniques:
\begin{itemize}
    \item \ul{Linear Techniques}: These methods assume that the data lies on or near a linear subspace of the original feature space. Examples include Principal Component Analysis (PCA) and Linear Discriminant Analysis (LDA).
    \item \ul{Non-Linear Techniques}: These methods can capture complex, non-linear relationships in the data. Examples include t-Distributed Stochastic Neighbor Embedding (t-SNE) and Uniform Manifold Approximation and Projection (UMAP).
\end{itemize}

\subsection{PCA (Principal Component Analysis)}

PCA is a widely used linear dimensionality reduction technique that transforms the original features into a new set of uncorrelated features called principal components. It works as follows:
\begin{enumerate}
    \item Standardize the data to have a mean of zero and a standard deviation of one.
    \item Compute the covariance matrix of the standardized data.
    \item Calculate the eigenvalues and eigenvectors of the covariance matrix.
    \item Sort the eigenvalues in descending order and select the top $k$ eigenvectors corresponding to the largest eigenvalues. These eigenvectors form the new feature space.
    \item Project the original data onto the new feature space to obtain the reduced-dimensional representation.
\end{enumerate}

The principal components capture the directions of maximum variance in the data, allowing for effective dimensionality reduction while retaining most of the information. The number of principal components to retain can be determined based on the desired level of explained variance. First components capture the most variance, while later components capture less.

\subsubsection{Eigenfaces}

Eigenfaces is a practical application of PCA in the field of computer vision, specifically for facial recognition. The principal components are called ``eigenfaces'', and when represented as images, they resemble ghostly faces. By projecting facial images onto the eigenface space, one can efficiently recognize and reconstruct faces based on their principal components.

\subsection{Factor Analysis (FA)}

Factor Analysis is another linear dimensionality reduction technique that models the observed variables as linear combinations of latent factors (basic structures of the data), plus some error terms. It assumes that the data follows a Gaussian distribution and aims to identify the underlying structure in the data. FA is particularly useful when the number of observed variables is large, and the eigenvalues and eigenvectors of the covariance matrix are computationally expensive to calculate.

\subsection{FastICA}

Fast Independent Component Analysis (FastICA) is an optimization algorithm for performing Independent Component Analysis (ICA). It is designed to find a linear transformation that maximizes the statistical independence of the components. FastICA is particularly effective for separating mixed signals, such as in blind source separation problems, where the goal is to recover original source signals from observed mixtures.

\subsection{SVD (Singular Value Decomposition)}

SVD is a general matrix factorization technique that decomposes a matrix into three matrices:
\begin{equation*}
    A = U \Sigma V^T
\end{equation*}
where:
\begin{itemize}
    \item $U$ is a rotation matrix.
    \item $\Sigma$ is a diagonal matrix containing the singular values, which represent the scaling factors of each component.
    \item $V^T$ is another rotation matrix.
\end{itemize}
SVD can be used for dimensionality reduction by retaining only the top $k$ singular values and their corresponding singular vectors, effectively capturing the most significant features of the data.

\subsection{t-SNE (t-Distributed Stochastic Neighbor Embedding)}

t-SNE is a non-linear dimensionality reduction technique that is particularly effective for visualizing high-dimensional data in 2D or 3D. As the main advantage, t-SNE focuses on preserving local structures in the data, revealing clusters and patterns that linear techniques might miss. It works by converting the high-dimensional distances between data points into probabilities and then minimizing the Kullback-Leibler divergence between the original and the reduced-dimensional representations.
\begin{enumerate}
    \item Construct a probability distribution over pairs of high-dimensional objects in such a way that similar objects have a higher probability of being picked, while dissimilar points have a lower probability.
    \item Define a similar probability distribution over the points in the low-dimensional map.
    \item Map the high-dimensional data to a low-dimensional space by minimizing the Kullback-Leibler divergence between the two distributions using gradient descent.
\end{enumerate}

\begin{observacion}
    The SNE algorithm, which is the predecessor of t-SNE, had a significant drawback in that it could not effectively handle the so-called ``crowding problem'', where points that are close together in high-dimensional space could end up being mapped far apart in the low-dimensional representation. t-SNE addresses this issue by using a Student's t-distribution to model the similarities in the low-dimensional space instead of a Gaussian distribution, which allows it to better preserve local structures and reveal clusters in the data.
\end{observacion}

\subsection{UMAP (Uniform Manifold Approximation and Projection)}
UMAP is an extension of t-SNE that is designed to be faster and more scalable while still preserving both:
\begin{itemize}
    \item \ul{Local Structure}: Like t-SNE, UMAP focuses on preserving the local relationships between data points, ensuring that similar points remain close together in the reduced-dimensional space.
    \item \ul{Global Structure}: UMAP also aims to preserve the global structure of the data, which allows it to maintain the overall shape and distribution of the data in the low-dimensional representation.
\end{itemize}


\section{Linear Regression}

Linear regression is the main core of this chapter, and it is a fundamental technique for modeling the relationship between a dependent variable ($y$) and one or more independent variables ($x$). The goal of linear regression is to find the best-fitting line (or hyperplane in higher dimensions) that minimizes the distance between the observed data points and the predicted values from the model. The basic form of a simple linear regression model can be expressed as:
\begin{equation*}
    y = \beta x + \epsilon
\end{equation*}

If $x$ is a vector of independent variables, then $\beta$ is a vector of coefficients, and $\beta_j$ represents the change in $y$ for a one-unit change in $x_j$, holding all other variables constant. The error term $\epsilon$ captures the variability in $y$ that cannot be explained by the linear relationship with $x$.

Some different measures of the error in linear regression include:
\begin{itemize}
    \item \ul{Mean Absolute Error (MAE)}: The average of the absolute differences between observed and predicted values.
    \begin{equation*}
        MAE = \frac{1}{n} \sum_{i=1}^{n} |y_i - \hat{y}_i|
    \end{equation*}
    \item \ul{Mean Squared Error (MSE)}: The average of the squared differences between observed and predicted values.
    \begin{equation*}
        MSE = \frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
    \end{equation*}
    \item \ul{Root Mean Squared Error (RMSE)}: The square root of the MSE, providing an error measure in the same units as the dependent variable.
    \begin{equation*}
        RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
    \end{equation*}
    \item \ul{Mean Absolute Percentage Error (MAPE)}: The average of the absolute percentage differences between observed and predicted values.
    \begin{equation*}
        MAPE = \frac{1}{n} \left(\sum_{i=1}^{n} \left| \frac{y_i - \hat{y}_i}{y_i} \right|\right)\cdot 100\%
    \end{equation*}
    \item \ul{R-squared ($R^2$)}: A statistical measure that represents the proportion of the variance in the dependent variable that is explained by the independent variables in the model.
    \begin{equation*}
        R^2 = 1 - \dfrac{\sum\limits_{i=1}^{n} (y_i - \hat{y}_i)^2}{\sum\limits_{i=1}^{n} (y_i - \bar{y})^2}
    \end{equation*}
    An $R^2$ value of 1 indicates that the model perfectly explains the variance in the dependent variable, while an $R^2$ value of 0 indicates that the model does not explain any of the variance.
\end{itemize}

In order to determine the best-fitting line, we can use different methods, such as:
\begin{itemize}
    \item \ul{Ordinary Least Squares (OLS)}: This method minimizes the sum of the squared differences between observed and predicted values, effectively minimizing the MSE. The loss function for OLS can be expressed as:
    \begin{equation*}
        L = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
    \end{equation*}

    \item \ul{Lasso Linear Regression (L1)}: If the dimensionality reduction of the data was not sufficient, overfitting can occur. Lasso regression is a regularization technique that can help mitigate overfitting by adding a penalty term to the loss function based on the absolute values of the coefficients. This may lead to some unimportant coefficients being reduced to zero, effectively performing \emph{feature selection} and {reducing the complexity of the model}. The loss function for Lasso can be expressed as:
    \begin{equation*}
        L = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^{p} |\beta_j|
    \end{equation*}
    where $\lambda$ is the regularization parameter that controls the strength of the penalty.

    \item \ul{Ridge Linear Regression (L2)}: Ridge regression is another regularization technique that adds a penalty term based on the squared values of the coefficients. Coefficients are \emph{shrunk towards zero} but not exactly zero, which can help to reduce the variance of the model without eliminating any features entirely. This method is particularly effective in cases where there is \emph{multicollinearity} among the independent variables, as it can help to reduce the variance of the coefficient estimates. The loss function for Ridge can be expressed as:
    \begin{equation*}
        L = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^{p} \beta_j^2
    \end{equation*}
    where $\lambda$ is the regularization parameter that controls the strength of the penalty.

    \item \ul{Elastic Net Linear Regression}: Elastic Net is a regularization technique that combines both Lasso and Ridge regression. It adds a penalty term that is a linear combination of the L1 and L2 penalties, allowing for both feature selection and coefficient shrinkage. The loss function for Elastic Net can be expressed as:
    \begin{equation*}
        L = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 + \lambda_1 \sum_{j=1}^{p} |\beta_j| + \lambda_2 \sum_{j=1}^{p} \beta_j^2
    \end{equation*}
    where $\lambda_1$ and $\lambda_2$ are the regularization parameters that control the strength of the L1 and L2 penalties, respectively.

    \begin{observacion}
        The choice of regularization technique and the tuning of the regularization parameters ($\lambda$) are crucial for achieving the best performance of the regression model. Cross-validation is often used to select the optimal values for these parameters, ensuring that the model generalizes well to unseen data. In that case, it is called Elastic Net Cross-Validation (ElasticNetCV).
    \end{observacion}
\end{itemize}

\section{Outlier detection}

Outliers are data points that deviate significantly from the majority of the data, and they can have a substantial impact on the performance of regression models. In order to work with outliers \emph{without having to detect them}, we can use robust regression techniques that are less sensitive to outliers. The main robust regression technique is RANSAC (Random Sample Consensus).
\begin{itemize}
    \item \ul{RANSAC (Random Sample Consensus)}: RANSAC is a robust regression algorithm that is designed to fit a model to data that may contain outliers. It works as follows:
    \begin{enumerate}
        \item Randomly select a subset of the data points and fit a model to this subset (using one of the previous regression techniques).
        \item Determine how many data points from the entire dataset fit the model within a certain tolerance (these points are called inliers).
        \item Repeat the process a specified number of times, and keep the model that has the highest number of inliers.
    \end{enumerate}

    RANSAC is particularly effective in situations where the dataset contains some outliers that could significantly affect the performance of traditional regression methods. However, some disadvantages of RANSAC include its sensitivity to the choice of parameters (such as the number of iterations and the tolerance level) and its computational cost, especially for large datasets.
\end{itemize}

However, there are also other methods for outlier detection that can be used in conjunction with regression models. They all produce a \emph{decision boundary} that separates inliers from outliers, and they can be used to identify and remove outliers from the dataset before fitting a regression model. In the following, we will discuss some of these methods.
\subsection{LOF (Local Outlier Factor)}

LOF is an unsupervised anomaly detection method that identifies outliers based on the local density of data points. It compares the local density of a point to the local densities of its neighbors, and points that have a significantly lower density than their neighbors are considered outliers. The algorithm works as follows:
\begin{enumerate}
    \item For each pair of data points $A, B$, calculate the distance $d(A, B)$ and the $k-\text{distance}(B)$, which is the distance to the $k$-th nearest neighbor of $B$. The reachability distance is defined as:
    \begin{equation*}
        \text{reachability\_distance}(A, B) = \max\{k-\text{distance}(B), d(A, B)\}
    \end{equation*}
    It represents the distance from $A$ to $B$ while considering the local density around $B$.

    \item For each data point $A$, calculate the local reachability density (LRD) as the inverse of the average reachability distance to its $k$ nearest neighbors:
    \begin{equation*}
        \text{LRD}(A) = \left( \frac{\sum\limits_{B \in N_k(A)} \text{reachability\_distance}(A, B)}{|N_k(A)|} \right)^{-1}
    \end{equation*}

    \item Finally, calculate the LOF score for each data point $A$ as the average ratio of the local reachability density of $A$ to the local reachability densities of its $k$ nearest neighbors:
    \begin{equation*}
        \text{LOF}(A) = \frac{1}{|N_k(A)|} \left( \sum_{B \in N_k(A)} \text{LRD}(B)\right) \cdot \dfrac{1}{\text{LRD}(A)}
    \end{equation*}
\end{enumerate}

Depending of the value of the LOF score, we can determine whether a data point is an outlier or not.
\begin{itemize}
    \item If $\text{LOF}(A) \approx 1$, then $A$ has a local density similar to its neighbors, and it is not considered an outlier.
    \item If $\text{LOF}(A) > 1$, then $A$ has a lower local density than its neighbors, and it is considered an outlier. The higher the LOF score, the more likely it is that $A$ is an outlier.
    \item If $\text{LOF}(A) < 1$, then $A$ has a higher local density than its neighbors, and it is considered an inlier.
\end{itemize}

\subsection{Isolation Forest}

Isolation Forest is an unsupervised anomaly detection algorithm that isolates anomalies by randomly partitioning the data using decision trees. The main idea behind Isolation Forest is that anomalies are easier to isolate than normal points, as they tend to be few and different from the majority of the data. The algorithm works as follows:
\begin{enumerate}
    \item Randomly select a feature and a split value to partition the data into two subsets.
    \item Repeat the partitioning process recursively until all data points are isolated (i.e., each point is in its own leaf node).
    \item The path length from the root to a leaf node is used to determine the anomaly score for each data point. Outliers tend to have shorter path lengths, as they are isolated more quickly than normal points.
\end{enumerate}

In order to have a more robust anomaly score, Isolation Forest typically builds an ensemble of trees (hence the name ``forest'') and averages the path lengths across all trees to compute the final anomaly score for each data point.



