\section{Introduction}

The Python part of this exercise is available in \myhref{https://github.com/LosDelDGIIM/LosDelDGIIM.github.io/blob/main/subjects/Erasmus-DUE/GKI/Exercises/Exercise00.ipynb}{this Jupyter Notebook}.

\subsection*{Intelligence Tests}

\begin{ejercicio}
    Turing Test vs. Lovelace Test: Which test focuses on creativity and which on simulating conversation? How do these two tests work?\\

    The Turing Test focuses on simulating conversation. It works by having a human evaluator engage in a natural language conversation with both a machine and a human without knowing which is which. If the evaluator cannot reliably distinguish the machine from the human, the machine is considered to have passed the Turing Test.

    The Lovelace Test, on the other hand, focuses on creativity. It requires an AI to create something original (like a piece of art, music, or a story) that cannot be attributed to any human creator. The AI must be able to explain how it created the work and why it is original, demonstrating a level of creativity that goes beyond mere imitation.
\end{ejercicio}

\begin{ejercicio}
    Why do many simple chatbots fail the Winograd Schema?\\

    Many simple chatbots fail the Winograd Schema because they lack the necessary contextual understanding and common sense reasoning required to correctly interpret the ambiguous sentences used in the test. The Winograd Schema consists of sentences that have a pronoun with an ambiguous antecedent, and the correct interpretation depends on understanding the context and the relationships between the entities involved. Simple chatbots often rely on pattern matching or statistical methods rather than true comprehension, which leads to incorrect answers when faced with such nuanced language challenges.
\end{ejercicio}

\begin{ejercicio}
    What is the main idea behind the Metzinger test for checking if an AI is self-aware, in simple terms? What is one thing that might make this test difficult to use for all types of AI?\\

    The main idea behind the Metzinger test is to check if an AI can understand and describe its own internal processes and experiences, which would indicate self-awareness. One difficulty with this test is that it relies on the AI's ability to communicate its thoughts and feelings, which may not be possible for all types of AI, especially those that do not have a natural language processing capability or a way to express their internal states effectively. Therefore, there is no universally applicable method to determine if an AI is self-aware, as it may depend on the specific design and capabilities of the AI in question.
\end{ejercicio}