\section{Neural Network}

The Python part of this exercise is available in \myhref{https://github.com/LosDelDGIIM/LosDelDGIIM.github.io/blob/main/subjects/Erasmus-DUE/GKI/Exercises/Exercise8.ipynb}{this Jupyter Notebook}.\\

\begin{ejercicio}
    Explain the role of activation functions in a neural network. Describe two common activation functions (e.g., ReLU, sigmoid, $\tanh$) and discuss a scenario where one might be preferred over the other.

    Activation functions introduce non-linearity into the neural network, allowing it to learn complex patterns in the data. Without activation functions, a neural network would essentially be a linear model, regardless of the number of layers.
    Two common activation functions are:
    \begin{itemize}
        \item \ul{ReLU (Rectified Linear Unit)}: Defined as $f(x) = \max(0, x)$, ReLU is computationally efficient and helps mitigate the vanishing gradient problem. It is often preferred in hidden layers of deep networks.
        \item \ul{Sigmoid}: Defined as $f(x) = \frac{1}{1 + e^{-x}}$, the sigmoid function maps input values to a range between 0 and 1. It is commonly used in the output layer for binary classification problems, as it can be interpreted as a probability.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    Describe the process of backpropagation during the training of a neural network.

    Backpropagation is a supervised learning algorithm used to train neural networks. It consists of several steps:
\begin{enumerate}
    \item \ul{Forward Pass}: The input data is passed through the network to compute the output.
    \item \ul{Loss Calculation}: The output is compared to the true labels using a loss function (e.g., Mean Absolute Error, Cross-Entropy Loss) to calculate the error.
    \item \ul{Backward Pass}: The error is propagated backward through the network to compute the gradients of the loss with respect to each weight in the network.
    \item \ul{Weight Update}: The weights are updated using an optimization algorithm (e.g., Gradient Descent) based on the computed gradients to minimize the loss.
\end{enumerate}
\end{ejercicio}

\begin{ejercicio}
    What is the purpose of a loss function in neural network training?

    A loss function is a mathematical function that quantifies the difference between the predicted output of the neural network and the true labels. The purpose of the loss function is to provide a measure of how well the model is performing, guiding the training process by indicating how much the model's predictions deviate from the actual values. During training, the optimization algorithm seeks to minimize the loss function, which in turn improves the model's performance on the training data.
\end{ejercicio}

\begin{ejercicio}
    Explain the difference between a classification task and a regression task.

    A classification task involves predicting a discrete label or category for each input instance. For example, classifying images of animals into categories such as ``cat'', ``dog'', or ``bird''. The output of a classification model is typically a probability distribution over the possible classes.

    A regression task, on the other hand, involves predicting a continuous value for each input instance. For example, predicting the price of a house based on its features (size, location, etc.). The output of a regression model is a single continuous value or a vector of continuous values.
\end{ejercicio}

\begin{ejercicio}
    Discuss the concepts of overfitting and underfitting in the context of neural network training. How can you identify these issues by analyzing the training and validation loss curves? What are two common techniques to mitigate overfitting?
    \begin{itemize}
        \item \ul{Overfitting}: Occurs when a model learns the training data too well, including its noise and outliers, resulting in poor generalization to new, unseen data. It can be identified by a low training loss and a high validation loss.
        \item \ul{Underfitting}: Occurs when a model is too simple to capture the underlying patterns in the data, resulting in poor performance on both the training and validation sets. It can be identified by high training and validation losses.
    \end{itemize}
    Two common techniques to mitigate overfitting are:
    \begin{itemize}
        \item \ul{Regularization}: Techniques such as L1 or L2 regularization add a penalty term to the loss function to discourage complex models and prevent overfitting.
        \item \ul{Dropout}: A technique that randomly drops a fraction of the neurons during training, which helps prevent the model from becoming too reliant on any particular set of features and encourages it to learn more robust representations.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    Why does a neural network need an activation function in the first place? What would happen if we completely omitted it?

    A neural network needs an activation function to introduce non-linearity into the model. Without activation functions, the network would only be able to learn linear relationships between the input and output, regardless of the number of layers. This would severely limit the model's capacity to capture complex patterns in the data. If we completely omitted activation functions, the network would effectively become a linear regression model, and it would not be able to solve problems that require learning non-linear decision boundaries, such as image classification or speech recognition.
\end{ejercicio}

\begin{ejercicio}
    What is Softmax? (Briefly explain its purpose and mathematical functionality).

    Softmax is an activation function that converts a vector of raw scores (logits) into a probability distribution over multiple classes. It is commonly used in the output layer of multi-class classification neural networks. The Softmax function takes as input a vector of real numbers and outputs a vector of values between 0 and 1 that sum to 1, representing the probabilities of each class. It includes a parameter $T$ (temperature) that can be used to control the ``"sharpness'' of the output distribution. The mathematical formula for Softmax is given by:
\[
\text{Softmax}(x_i) = \frac{e^{\nicefrac{x_i}{T}}}{\sum\limits_{j} e^{\nicefrac{x_j}{T}}}
\]
    where $j$ indexes over all classes, $x_i$ is the input score for class $i$, and $T$ is the temperature parameter. A lower temperature results in a more confident (peaked) distribution, while a higher temperature produces a softer (more uniform) distribution.
\end{ejercicio}

\begin{ejercicio}
    What exactly does the ``Loss'' measure during training, and why does the optimizer try to make this value as small as possible?

    The ``Loss'' measures the discrepancy between the predicted output of the neural network and the true labels of the training data. It quantifies how well the model is performing on the training examples. A lower loss indicates that the model's predictions are closer to the actual values, while a higher loss indicates greater error in the predictions.
    The optimizer tries to minimize the loss value because doing so improves the model's performance. By reducing the loss, the model learns to make more accurate predictions, which is the ultimate goal of training a neural network. A smaller loss value generally corresponds to better generalization to unseen data, leading to improved performance on real-world tasks.
\end{ejercicio}

\begin{ejercicio}
    Name 3 common optimizers that can be used when training neural networks.
    % // TODO: Hacer
\end{ejercicio}

\begin{ejercicio}
    % // TODO: Corregir
    Provide the mathematical formula for the Cross-Entropy loss function (for a single sample in multi-class classification) and explain its individual components.

    The Cross-Entropy loss function for a single sample in multi-class classification is given by the formula:
\[
L = -\sum_{i=1}^{C} y_i \log(p_i)
\]
where:
\begin{itemize}
    \item $C$ is the number of classes.
    \item $y_i$ is the true label for class $i$, which is 1 if the sample belongs to class $i$ and 0 otherwise (one-hot encoded).
    \item $p_i$ is the predicted probability for class $i$, obtained from the Softmax output of the neural network.
\end{itemize}
\end{ejercicio}


\begin{ejercicio}[Bonus]
    Explain the concept of a batch size in the context of training a neural network with a DataLoader. What are the potential benefits and drawbacks of using a smaller versus a larger batch size?
    % // TODO: Hacer
    \begin{comment}
    The batch size is the number of training examples used in one forward and backward pass of the training process before the network's weights are updated. A `DataLoader` divides the entire training dataset into smaller batches.

    * **Benefits of a smaller batch size:**
        * **Regularizing effect:** Smaller batches provide a noisier estimate of the gradient, which can help the model escape local minima and improve generalization (similar to stochastic optimization).
        * **Lower memory requirements:** Smaller batches require less memory on the GPU or CPU.
        * **More frequent weight updates:** The model's weights are updated more frequently, which can lead to faster learning in some cases.

    * **Drawbacks of a smaller batch size:**
        * **Slower training per epoch:** More weight updates mean more computations per epoch.
        * **Noisier gradient estimates:** The noisy gradient can make the training process more unstable.

    * **Benefits of a larger batch size:**
        * **Faster training per epoch:** Fewer weight updates mean fewer computations per epoch.
        * **More stable gradient estimates:** The gradient is estimated more accurately as it's based on more examples.
        * **More efficient hardware utilization:** Can better utilize parallel processing on GPUs.

    * **Drawbacks of a larger batch size:**
        * **Potential for poorer generalization:** The model might get stuck in sharp minima of the loss landscape and generalize worse to new data.
        * **Higher memory requirements:** Larger batches require more memory.
    \end{comment}
\end{ejercicio}

\begin{ejercicio}[Bonus]
    What is regularization in the context of neural networks? Describe one common regularization technique (e.g., dropout) and explain how it can help improve the generalization ability of a model.
    % // TODO: Hacer
    \begin{comment}
    Regularization refers to techniques used to prevent overfitting in machine learning models, including neural networks. They add extra constraints or information to the learning algorithm to prevent the model from becoming too complex and fitting the training data too closely. The goal is to improve the model's ability to generalize to unseen data.

    * **Dropout:** Dropout is a regularization technique where, during each training iteration, a randomly selected proportion of neurons in a layer are set to zero. This means that these neurons are not activated in the forward pass and do not participate in the backward pass.

        * **How it helps improve generalization:**
            * **Prevents co-adaptation of features:** Because neurons are randomly dropped out, the remaining neurons cannot rely too heavily on specific features of the training data that might not be present in new data. They are forced to learn more robust features that are useful across different examples.
            * **Ensemble effect:** Each dropout mask (the random selection of dropped-out neurons) effectively creates a different "sub-network." Training with dropout can be seen as training an ensemble of many such sub-networks, and their predictions are implicitly averaged during testing. This often leads to more robust and better-generalizing performance.
            * **Reduces reliance on individual neurons:** The network becomes more resilient to the failure of individual neurons, as the information has to be distributed across multiple neurons.
    \end{comment}
\end{ejercicio}