\section{Data Processing / Cleaning}

The Python part of this exercise is available in \myhref{https://github.com/LosDelDGIIM/LosDelDGIIM.github.io/blob/main/subjects/Erasmus-DUE/GKI/Exercises/Exercise1.ipynb}{this Jupyter Notebook}.

\subsection*{Data Filtering and Transforming}

\begin{ejercicio}
    Imagine you have a list of all the students in a school. You only want to look at the students in the 5th grade (data filtering). Then, you want to calculate the average age of those 5th graders (data transforming). Explain why you need both steps.\\

    We first need to filter the data to focus only on the students in the 5th grade, as we are specifically interested in that group. This step allows us to narrow down our dataset to the relevant subset of students. After filtering, we need to transform the data by calculating the average age of those 5th graders. This transformation step is necessary to derive meaningful insights from the filtered data, as it provides us with a summary statistic (the average age) that can help us understand the characteristics of the 5th-grade students. Without filtering, we would be calculating the average age of all students, which would not give us the specific information we are looking for about the 5th graders.
\end{ejercicio}

\begin{ejercicio}
    You have a list of toys that were sold in a store, including the toy's name and the price it sold for.
    \begin{enumerate}
        \item Give an example of a question you could answer by only looking at toys that cost more than $20$ (data filtering).
        
        A question that could be answered by data filtering would be: ``Which toys that cost more than 20 euros were sold in the last month?'' (Here, we filter the data by price greater than 20 euros and potentially by the sales date).
        \item Give an example of something you could calculate or change about the toy prices (data transforming).
        
        An example of data transformation would be calculating the average selling price of all toys or converting all prices from euros to another currency (e.g., US dollars) based on a specific exchange rate. Another transformation could be creating a new column indicating whether a toy is classified as ``cheap'' (e.g., $<$ 15 euros), ``mid-priced'' (e.g., 15 euros $-$ 30 euros), or ``expensive'' ($>$ 30 euros) based on the selling price.
    \end{enumerate}
\end{ejercicio}

\subsection*{Data Understanding}

\begin{ejercicio}
    What is the difference between quantitative (numerical) and qualitative (categorical) data?\\

    Quantitative data refers to numerical values that can be measured and quantified, such as height, weight, or age. Qualitative data, on the other hand, refers to categorical values that describe characteristics or attributes, such as color or gender. Quantitative data can be analyzed using mathematical operations, while qualitative data is typically analyzed using descriptive statistics or categorization.
\end{ejercicio}

\begin{ejercicio}
    What does the abbreviation \verb|NaN| mean in a dataset?

    \verb|NaN| stands for ``Not a Number'' and is used in datasets to represent missing or undefined values. It is also used to indicate that a value is not a valid number, such as the result of an undefined mathematical operation (e.g., division by zero). In data analysis, \verb|NaN| values need to be handled appropriately, either by filling them with a specific value or dropping them.
\end{ejercicio}

\begin{ejercicio}
    Briefly explain the difference between the mean and the median. Which value is more sensitive to outliers?

    The mean is the average of a set of numbers, calculated by summing all the values and dividing by the count of values. The median is the middle value in a sorted list of numbers. The mean is more sensitive to outliers because it can be significantly affected by extremely high or low values, while the median is less affected as it only considers the middle value(s) and not the magnitude of all values.
\end{ejercicio}

\begin{ejercicio}
    Calculate the quartiles for the following dataset: [10, 9, 1, 4, 3, 2, 5, 6, 6, 6, 3, 9].

    First, we need to sort the dataset:
    \begin{equation*}
    [1, 2, 3, 3, 4, 5, 6, 6, 6, 9, 9, 10]
    \end{equation*}

    The quartiles are then:
    \begin{align*}
        Q2 &= \dfrac{6 + 5}{2} = 5.5\\
        Q1 &= \dfrac{3 + 3}{2} = 3\\
        Q3 &= \dfrac{6 + 9}{2} = 7.5
    \end{align*}
\end{ejercicio}

\subsection*{Data Cleaning}

\begin{ejercicio}
    Name two reasons why data is often dirty in the real world.\\

    Data can be dirty due to:
    \begin{itemize}
        \item Human error during data entry, which can lead to typos, incorrect values, or missing data.
        \item Inconsistent data formats or structures, such as different date formats or varying units of measurement across datasets.
    \end{itemize}
\end{ejercicio}

\begin{ejercicio}
    What happens when you apply the \verb|dropna()| function to a DataFrame?

    The \verb|dropna()| function removes any rows (or columns, if specified with the \verb|axis=1| parameter) that contain missing values (\verb|NaN|) from the DataFrame. The original DataFrame is not modified unless the \verb|inplace=True| parameter is used.
    
    This is a common data cleaning step to ensure that the dataset does not contain incomplete records that could affect analysis or modeling.
\end{ejercicio}

\begin{ejercicio}
    Why do you scale (normalize) numerical data before feeding it into a machine learning model?\\

    Scaling (normalizing) numerical data is important because many machine learning algorithms are sensitive to the scale of the input features. If the features have different ranges, the algorithm may give more weight to those with larger values, leading to biased results. Scaling helps to ensure that all features contribute equally to the model's learning process and can improve the convergence speed and performance of the model.
\end{ejercicio}

\begin{ejercicio}
    Name two possible encodings and explain them using an example.\\

    They are explained in the page~\pageref{item:encoding}.
\end{ejercicio}

\begin{ejercicio}
    Describe three common challenges encountered during data preparation and explain a technique to mitigate each challenge.
    \begin{itemize}
        \item \ul{Missing Data}: This can be mitigated by using techniques such as imputation (filling in missing values with the mean, median, or mode) or by dropping rows/columns with a high percentage of missing values.
        \item \ul{Outliers}: Outliers can skew the results of analysis and modeling. They can be mitigated by using methods such IQR to identify and either remove or transform outliers.
        \item \ul{Inconsistent Data}: Inconsistent data can arise from different formats or units. This can be mitigated by standardizing the data format (e.g., converting all dates to a common format) and ensuring that all measurements are in the same units (e.g., converting all weights to kilograms). 
    \end{itemize}
\end{ejercicio}


\begin{ejercicio}
    Explain the difference between data cleaning and data transformation. Provide an example of each.
    \begin{description}
        \item[Data Cleaning] Data cleaning involves identifying and correcting errors or inconsistencies in the dataset to ensure that it is accurate and reliable. For example, removing duplicate entries from a customer database or filling in missing values in a sales dataset.
        \item[Data Transformation] Data transformation involves converting data from one format or structure to another to make it suitable for analysis or modeling. For example, normalizing numerical features to a common scale or encoding categorical variables into numerical format using one-hot encoding. 
    \end{description}
\end{ejercicio}

\begin{ejercicio}
    Why is it important to handle missing data appropriately? Discuss three different strategies for dealing with missing values and their potential drawbacks.

    Handling missing data appropriately is crucial because it can significantly impact the results of analysis and modeling. If not addressed, missing data can lead to biased estimates, reduced statistical power, and inaccurate predictions. Three strategies for dealing with missing values are:
\begin{itemize}
    \item \ul{Imputation}: Filling in missing values with a specific value (e.g., mean, median, mode) can help maintain the dataset's size and structure. However, it can introduce bias if the imputed values do not accurately reflect the underlying distribution of the data.
    \item \ul{Deletion}: Removing rows or columns with missing values can be simple but may lead to loss of information and reduced sample size.
    \item \ul{Model-based Imputation}: Using statistical models to predict and fill in missing values based on other available data. This approach can be more accurate but is computationally intensive and may introduce additional complexity in model interpretation.
\end{itemize}
\end{ejercicio}


\begin{ejercicio}
    What are outliers in a dataset and why can they be problematic? Describe three methods for detecting outliers.

    Outliers are data points that significantly differ from the majority of the data. They can be problematic because they can skew the results of analysis and modeling, leading to inaccurate conclusions. Three methods for detecting outliers are:
\begin{itemize}
    \item \ul{Z-score}: This method identifies outliers by calculating the number of standard deviations a data point is from the mean. A common threshold is a Z-score of 3 or -3.
    \item \ul{Interquartile Range (IQR)}: This method identifies outliers by calculating the IQR (the difference between the third quartile and the first quartile) and defining outliers as data points that fall below $Q1 - 1.5 \cdot IQR$ or above $Q3 + 1.5 \cdot IQR$.
    \item \ul{Visual Methods}: Box plots and scatter plots can visually reveal outliers in the data, allowing for easy identification of data points that deviate from the overall pattern.
\end{itemize}
\end{ejercicio}


\begin{ejercicio}
    Explain the concept of data normalization or standardization. When would you prefer one over the other, and why is this a crucial step in many machine learning workflows?

    Data normalization (scaling to a range) and standardization (scaling to have a mean of 0 and a standard deviation of 1) are techniques used to rescale numerical features. Normalization is preferred when the data does not follow a Gaussian distribution and when the algorithm does not assume any distribution. Standardization is preferred when the data follows a Gaussian distribution or when the algorithm assumes that the data is normally distributed. This step is crucial in machine learning workflows because it ensures that all features contribute equally to the model's learning process and can improve the convergence speed and performance of the model.
\end{ejercicio}