\chapter{Applications of Complex Numbers}

Complex numbers have a wide range of applications in various fields of science and engineering. In the following sections, we will explore some of the most common ones.

\section{Harmonic Oscillations}

Oscillations are a fundamental phenomenon in physics, and thousands of systems can be modeled using them, starting from the simple pendulum to the vibrations of a guitar string. Harmonic oscillations are a specific type of oscillation that can be described using a sine\footnote{Cosine functions can also be used, as they are just phase-shifted versions of sine functions.} function. The general form of a harmonic oscillation can be expressed as:
\begin{equation}
f(t) = A \sin(\omega t + \varphi)
\label{eq:harmonic_oscillation}
\end{equation}
where:
\begin{itemize}
    \item $A$ is the amplitude of the oscillation, representing the maximum displacement from the equilibrium position.
    \item $\omega$ is the angular frequency, which determines how many oscillations occur in a unit of time.
    \item $\varphi$ is the phase shift, which determines the initial position of the oscillation at time $t=0$.
    \item $t$ is the time variable.
    \item $f(t)$ is the value of the oscillation at time $t$.
\end{itemize}

In order to represent harmonic oscillations using complex numbers, the following complex function can be used, where the underline usually denotes a complex number:
\begin{align*}
\ul{z}(t) &= A e^{i(\omega t + \varphi)}
\\&= A \left( \cos(\omega t + \varphi) + i \sin(\omega t + \varphi) \right)
\end{align*}

Therefore, the harmonic oscillation described in Equation~\ref{eq:harmonic_oscillation} can be represented as $f(t)=\Im(\ul{z}(t))$, where $\Im$ denotes the imaginary part of the complex function. This representation allows us to analyze and manipulate harmonic oscillations just using complex numbers, which can simplify calculations and provide deeper insights into the behavior of oscillatory systems.

One additional advantage of using complex numbers is that the dependency on the time can be split. The number $\ul{z}$ can be written as:
\begin{equation*}
    \ul{z}(t) = A e^{i\varphi} e^{i\omega t}
\end{equation*}
where $A e^{i\varphi}$ is a complex constant that contains the amplitude and phase shift, while $e^{i\omega t}$ contains the time dependency. The value $A e^{i\varphi}$ is often referred to as the \textit{complex amplitude} $\ul{A}$, and $|\ul{A}|=A$ is the amplitude of the oscillation. This separation allows us to analyze the oscillation in terms of its amplitude and phase independently from its time evolution.

\section{Alternating Current Circuits}

Alternating current (AC) circuits are electrical circuits that operate with an alternating voltage or current. An alternating voltage can be represented as a sinusoidal function of time such as~\ref{eq:ac_voltage}, that generates an alternating current such as~\ref{eq:ac_current}:
\begin{align}
v(t) &= \hat{v} \sin(\omega t + \varphi_v)
\label{eq:ac_voltage}
\\i(t) &= \hat{i} \sin(\omega t + \varphi_i)
\label{eq:ac_current}
\end{align}
where $\hat{v}$ and $\hat{i}$ are the peak voltage and current, $\omega$ is the angular frequency, and $\varphi_v$ and $\varphi_i$ are the phase shifts of the voltage and current, respectively. In AC circuits, an osciloscope can be used to visualize the voltage and current waveforms, which often appear as sinusoidal waves. If a multimeter is used to measure the voltage and current, it can't just provide the average value (as the positive and negative halves of the wave cancel each other out), but it can provide the root mean square (RMS) value, which is a measure of the effective voltage or current.
\begin{definicion}[Root Mean Square]
The root mean square (RMS) value of a periodic function $f(t)$ is defined as:
\begin{equation*}
\text{RMS}(f) = \sqrt{\frac{1}{T} \int_0^T f(t)^2 dt}
\end{equation*}
where $T$ is the period of the function.
\end{definicion}
\begin{prop}
The RMS value of a sinusoidal function $f(t) = A \sin(\omega t + \varphi)$ is given by:
\begin{equation*}
\text{RMS}(f) = \frac{A}{\sqrt{2}}
\end{equation*}
\end{prop}
\begin{proof}
\begin{align*}
\text{RMS}(f) &= \sqrt{\frac{1}{T} \int_0^T A^2 \sin^2(\omega t + \varphi) dt}
= A \sqrt{\frac{1}{T} \int_0^T \sin^2(\omega t + \varphi) dt}
=\\&= A \sqrt{\frac{1}{T} \int_0^T \frac{1 - \cos(2\omega t + 2\varphi)}{2} dt}
= A \sqrt{\frac{1}{T} \left[ \frac{t}{2} - \frac{\sin(2\omega t + 2\varphi)}{4\omega} \right]_0^T}
\AstIg\\&\AstIg A \sqrt{\frac{1}{T} \left( \frac{T}{2} - 0 \right)}
= A \sqrt{\frac{1}{2}} = \frac{A}{\sqrt{2}}
\end{align*}
where in $(\ast)$ we used that $T$ is the period of the function.
\end{proof}

Therefore, the RMS values read by a multimeter for the voltage (usually denoted as $V$) and current ($I$) in an AC circuit should be multiplied by $\sqrt{2}$ to obtain the peak values $\hat{v}$ and $\hat{i}$, respectively. As happened with harmonic oscillations, complex numbers can be used to represent the voltage and current in AC circuits. The voltage can be represented as the imaginary part of a complex function:
\begin{align*}
\ul{v}(t) &= \hat{v} e^{j(\omega t + \varphi_v)}
= \hat{v}e^{j\varphi_v} e^{j\omega t}
= \ul{\hat{v}} e^{j\omega t}
\end{align*}
where $\ul{\hat{v}} = \hat{v}e^{j\varphi_v}$ is the complex amplitude of the voltage. The value $\ul{V}=\sqrt{2}\ul{\hat{v}}$ is often referred to as the \textit{complex RMS voltage}.

\subsection{Ohm's Law and Impedance}

Ohm's Law is a fundamental principle in electrical engineering that relates the voltage ($V$), current ($I$), and resistance ($R$) in a DC circuit:
\begin{equation}
V = IR
\label{eq:ohms_law}
\end{equation}

In AC circuits, a new concept called \textit{impedance} ($\ul{Z}$) is introduced to generalize Ohm's Law.
\begin{definicion}[Impedance]
The impedance of an AC circuit is a complex quantity that relates the voltage and current in the circuit, and is defined as:
\begin{equation*}
\ul{Z} = \frac{\ul{v}(t)}{\ul{i}(t)}
\end{equation*}
\end{definicion}

The general form of Ohm's Law for AC circuits can be expressed as:
\begin{align*}
\ul{Z} &= \frac{\ul{v}(t)}{\ul{i}(t)}
= \frac{\ul{\hat{v}} e^{j\omega t}}{\ul{\hat{i}} e^{j\omega t}}
= \frac{\ul{\hat{v}}}{\ul{\hat{i}}}
= \frac{\hat{v}e^{j\varphi_v}}{\hat{i}e^{j\varphi_i}}
= \frac{\hat{v}}{\hat{i}} e^{j(\varphi_v - \varphi_i)}
= \frac{V}{I} e^{j(\varphi_v - \varphi_i)}
\end{align*}
As we can observe, the impedance $\ul{Z}$ is a complex number that does not depend on time. It is usually divided into its real part, called \textit{resistance} ($R$), and its imaginary part, called \textit{reactance} ($X$). The impedance takes into account the effects of three types of circuit elements:
\begin{itemize}
    \item Resistors, which have a resistance $R$ measured in ohms ($\Omega$). The follow equation is always true for resistors:
    \begin{equation*}
        v(t) = R\cdot i(t)
    \end{equation*}

    As we can see, $\ul{Z}_R = R$. In addition, the voltage and current have the same phase and angular frequency, but the amplitude is different.
    \item Inductors, which have an inductance $L$ measured in henries ($H$). The follow equation is always true for inductors:
    \begin{equation*}
        \ul{v}(t) = L \frac{d\ul{i}(t)}{dt}
    \end{equation*}

    Operating the derivative, we can obtain:
    \begin{align*}
        \ul{v}(t) &= L \frac{d\ul{i}(t)}{dt}
        = L \frac{d}{dt} \left( \ul{\hat{i}} e^{j\omega t} \right)
        = j\omega L \ul{\hat{i}} e^{j\omega t}
        = j\omega L \ul{i}(t)
        \Longrightarrow
        \ul{Z}_L = j\omega L
    \end{align*}

    Taking into account the definition of impedance, we can conclude that $\ul{Z}_L =~j\omega L$. In addition, the voltage and current have the same angular frequency, but there is a phase shift of $\nicefrac{-\pi}{2}$ between them, and the amplitude is different.
    \item Capacitors, which have a capacitance $C$ measured in farads ($F$).
    
    The follow equation is always true for capacitors:
    \begin{equation*}
        \ul{i}(t) = C \frac{d\ul{v}(t)}{dt}
    \end{equation*}

    Operating the derivative, we can obtain:
    \begin{align*}
        \ul{i}(t) &= C \frac{d\ul{v}(t)}{dt}
        = C \frac{d}{dt} \left( \ul{\hat{v}} e^{j\omega t} \right)
        = j\omega C \ul{\hat{v}} e^{j\omega t}
        = j\omega C \ul{v}(t)
        \Longrightarrow
        \ul{Z}_C = \frac{1}{j\omega C}
    \end{align*}

    Taking into account the definition of impedance, we can conclude that $\ul{Z}_C=~\frac{1}{j\omega C}$. In addition, the voltage and current have the same angular frequency, but there is a phase shift of $\nicefrac{\pi}{2}$ between them, and the amplitude is different.
\end{itemize}


``Solving'' an AC circuit means calculating the voltage, current, and impedance in the circuit. In order to do so, first of all the equivalent impedance of the circuit must be calculated using the rules for series and parallel connections of impedances.
\begin{itemize}
    \item Serial connections of impedances are impedences that are connected end-to-end, so the same current flows through each impedance. The equivalent impedance of a series connection is given by the sum of the individual impedances:
    \begin{equation*}
        \ul{Z}_{\text{series}} = \sum_{k=1}^n \ul{Z}_k
    \end{equation*}
    \item Parallel connections of impedances are impedences that are connected across the same two points, so the same voltage is applied across each impedance. The equivalent impedance of a parallel connection is given by the reciprocal of the sum of the reciprocals of the individual impedances:
    \begin{equation*}
        \dfrac{1}{\ul{Z}_{\text{parallel}}} = \sum_{k=1}^n \frac{1}{\ul{Z}_k}
    \end{equation*}
\end{itemize}

After that, and appart from the Ohm's Law, the Kirchhoff's circuit laws can be used to solve the circuit.
\begin{itemize}
    \item Kirchhoff's Current Law (KCL) states that the total current entering a junction in a circuit must equal the total current leaving the junction. Mathematically, it can be expressed as:
    \begin{equation*}
        \sum_{k=1}^n I_k = 0
    \end{equation*}
    where $I_k$ is each current entering or leaving the junction, with currents entering the junction considered positive and currents leaving the junction considered negative.
    \item Kirchhoff's Voltage Law (KVL) states that the total voltage around any closed loop in a circuit must equal zero. Mathematically, it can be expressed as:
    \begin{equation*}
        \sum_{k=1}^n V_k = 0
    \end{equation*}
    where $V_k$ is the voltage difference across each element in the loop, following a consistent direction (e.g., clockwise or counterclockwise).
\end{itemize}